---
title: "Lotties | Installation and server setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lottie)
```

[`{lottie}`][lottie] is a R package with a [Shiny][shiny] app for capturing data from a long-running field study of Long
Tailed Tits in the Rivelin Valley on the edge of the city of Sheffield, UK.

## Installation

You can install the package directly from GitHub using [`{pak}`][pak], if you don't have `{pak}` installed install it first).

```shell
install.packages("pak")
pak::pkg_install("Morinay-lab/lotties")
```

## Launching

To launch the Shiny App locally you can run the following...

```R
shiny::runApp(system.file("shiny", package = "lotties"))
```

This should launch a browser pointing to a locally running instance of the application. The database will be in memory
and is _not_ permanent.

### Persistence

By default the App is set to run in memory and so no historical data nor data entered persists. When the application is
killed all entered data will be lost. Included in the package under `inst/xls` are various historical data files which
have been cleaned and added to a [SQLite][sqlite3] which resides under `inst/sqlite`. By modifying `testing <- FALSE` in
the `inst/shiny/server.R` file then the database with the historical data in the `inst/sqlite/` directory will be
loaded.

### Data Entry

There are a number of sections for data to be entered. It is expected that they will be completed in the order in which
they appear on the "accordian" on the left, i.e. _GPS > Observation Metadata > Flock Data > Individual Data >
Interactions Data_. Doing so means that the `Flock ID` entered under the _Flock Data_ will then be used to populate the
pull-down menus in subsequent sections (_Individual Data_ and _Interactions Data_).

## Deployment

To make the Shiny App available more widely the Morinay Lab have comissioned a Virtual Machine (VM) from University of
Sheffield IT Services.

This is a GNU/Linux machine running Ubuntu Linux and users can connect to the Shiny App that is running from any
computer on the University network or whilst connected to the Universities Virtual Private Network (VPN).

### User account

The `lotties` account has been created to run the service, once you have SSH'd into the machine you can switch
to this account using

```shell
sudo su lotties
cd
```

### Get the repository

The `lotties` account has a clone of the repository under `~/lotties`. You can update it by changing directory and
pulling.
```shell
cd ~/lotties
git pull
```

If it is not cloned, for example the service is being deployed elsewhere, you can clone it using

```
cd
git clone https://github.com/Morniay-Lab/lotties.git
```

### Launching the Service

A [systemd service module][systemd] has been added to facilitate starting and stopping the service. It has been setup to
be run as the `lotties` user on the account so after you have SSH'd to the machine you should switch to this account
(see above) and you can start

```shell
systemctl restart lotties.service
```

You can check the status with

```shell
systemctl status lotties.service
```

### Updating

The `lotties` package is a work in progress and new features may be added. To update these you should..

1. Switch to the `lotties` user account.
2. Navigate to `~/lotties/` git repository.
3. Get the latest changes to the `main` branch with `git switch main && git pull`
4. Edit `inst/shiny/server.R` and set `testing <- FALSE` so the SQLite database is used rather than running in memory.
5. Restart the service with `sudo systemctl restart lotties.service`

It is possible that you may wish to test features on a branch in which case you can replace step 3 with...

3. Switch to branch to be tested with `git switch <branch>`

...although ideally all branches should have been tested locally in memory and changes merged to `main` for deployment.

## Virtual Machine Administration

To connect to the virtual machine `lotties.shef.ac.uk` you will need to have an account created for you. For security
it has been configured to only allow connections using SSH authentication.

### Key Generation

#### GNU/Linux and OSX
If you are using GNU/Linux or OSX then you
can create SSH keys using the following commands (**NB** - It is recommend to use [Ed25519 Twisted Edward curve keys][ed25519])

```shell
ssh-keygen -o -a 100 -t ed25519
```

By default you should be prompted for a secure password.

#### Windows

Windows users should follow the instructions on [Using PuTTYgen on Windows to generate SSH key pairs][puttySSH].

### Uploading Keys

Share your `id_25519.pub` (the public part of your key) with the administrator who will setup your account and place the
public key under your `/home/<username>/.ssh/id_ed25519.pub`

### R setup

The University of Sheffield Virtual Machines run Ubuntu. We can leverage [r2u][r2u] to install binaries of all packages
which speeds up installation. The [Usage and setup](https://github.com/eddelbuettel/r2u#usage-and-setup) instructions
should be followed. This will, if you include _Step 3 : Ensure you have current R binaries_, install the latest
version of R. Once complete you can check the version of R

```
R --version
```


#### Package install

Switch to the `lotties` user to clone and install (or update) the `lotties` package

```
sudo su lotties
```

Start R and install the `devtools` package and use it to install the local clone of the ~lotties~ package along with dependencies.

```r
R
> install.packages("devtools")
> devtools::install("./", dependencies = TRUE)
```

If any packages fail then install them globally using the `r2u` built packages.

```r
> install.packages(c("patrick", "RSQLite", "ggplot2", "scales", "tidyr", "viridisLite", "watcher"))
> devtools::install("./", dependencies = TRUE)

-- pak::local_install_deps() ---------------------------------------------------------------------------------------

> Package library at /home/lotties/R/x86_64-pc-linux-gnu-library/4.6, /usr/local/lib/R/site-library, /usr/lib/R/sitebrary, and /usr/lib/R/library.
> The package (0 B) is cached.
x Missing 3 system packages. You'll probably need to install them manually:
+ cmake        - fs
+ libuv1-dev   - fs
+ libxml2-dev  - xml2
i No downloads are needed
v 86 deps: kept 86 [2.1s]
-- R CMD build -----------------------------------------------------------------------------------------------------
v  checking for file '/home/lotties/lotties/DESCRIPTION' ...
-  preparing 'lottie':
v  checking DESCRIPTION meta-information ...
-  checking for LF line-endings in source and make files and shell scripts
-  checking for empty or unneeded directories
     NB: this package now depends on R (>= 4.1.0)
     WARNING: Added dependency on R >= 4.1.0 because package code uses the
     pipe |> or function shorthand \(...) syntax added in R 4.1.0.
     File(s) using such syntax:
       'utils.R'
-  building 'lottie_0.0.0.9000.tar.gz'

-- R CMD INSTALL ---------------------------------------------------------------------------------------------------
Running /usr/lib/R/bin/R CMD INSTALL /tmp/RtmpKnTcHK/lottie_0.0.0.9000.tar.gz \
  --install-tests
* installing to library '/home/lotties/R/x86_64-pc-linux-gnu-library/4.6'
* installing *source* package 'lottie' ...
** this is package 'lottie' version '0.0.0.9000'
** using staged installation
** R
** inst
** tests
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** installing vignettes
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (lottie)
>
```

You can now launch the application, but you need to know what the IP address of the machine is. You can find this out
using either `ifconfig` or the newer `ip a`. The call can be made from a plain shell, or you can do it from the running
R session by using the `system()` function to make a system call.

```r
> system("ip a")


1: lo: <LOOPBACK,UP,LOWER_UP> mtu 65536 qdisc noqueue state UNKNOWN group default qlen 1000
    link/loopback 00:00:00:00:00:00 brd 00:00:00:00:00:00
    inet 127.0.0.1/8 scope host lo
       valid_lft forever preferred_lft forever
    inet6 ::1/128 scope host noprefixroute
       valid_lft forever preferred_lft forever
2: ens192: <BROADCAST,MULTICAST,UP,LOWER_UP> mtu 1500 qdisc mq state UP group default qlen 1000
    link/ether 00:50:56:b6:bd:a8 brd ff:ff:ff:ff:ff:ff
    altname enp11s0
    inet 123.456.178.9/24 brd 143.167.249.255 scope global ens192
       valid_lft forever preferred_lft forever
    inet6 fe80::250:56ff:feb6:bda8/64 scope link
       valid_lft forever preferred_lft forever
```

Here the network device is `ens192` and the IP address associated with it is `123.456.178.9`. We therefore load the
~lotties~ package and start the
Shiny App, pointing it to the location of the Shiny `server.R` and `ui.R` (within the `inst/shiny/` directory) and use the IP address...

```r
> devtools::load_all()
> shiny::runApp("inst/shiny", host = "123.456.178.9")
```

The port used will be `3781` as this is defined in the packages `.Rprofile` file. For convenience you can add the host
to this file too which will negate the need to explicitly provide the `host` parameter when calling `shiny::runApp()`
but since this is contingent on the computer it is being run on it is _not_ included in the version in this repository.

```
## Shiny options
options(shiny.autoreload = TRUE,
        shiny.port = 3781,
        shiny.host = "123.456.178.9",
        shiny.trace = TRUE)
```


# Appendix

## Systemd Service module

This is based on the [Using systemd to Run Shiny Apps][systemd_shiny] article ([archived][systemd_shiny_archive]) but
with a few tweaks to run it as user. It is saved in the `/etc/systemd/user/lotties.service` file.

```
[Unit]
Description=Lotties Shiny App
After=network.target
StartLimitIntervalSec=0

[Service]
Type=simple
Restart=always
RestartSec=1
User=<username>
ExecStart=nohup Rscript -e "devtools::load_all(path='/home/lotties/lotties/')" -e "shiny::runApp('/home/lotties/lotties/inst/shiny', port = ####, host = '###.###.###.###')" > /home/lotties/log/lotties.log 2>&1 &

[Install]
WantedBy=multi-user.target
```

**NB** - This should log output to `/home/lotties/log/lotties.log` but it isn't currently redirecting. Instead use
`journalctl -u lotties.service --since today` to see output from the service from the current day.

### Enable the service

If setting up on a new system copy the above to `/etc/systemd/user/lotties.service` and then as `root` enable the
service with...

```shell
systemctl enable --now lotties.service
```

[ed25519]: https://blog.g3rt.nl/upgrade-your-ssh-keys.html
[pak]: https://pak.r-lib.org/
[lottie]: https://github.com/Morinay-lab/lotties
[puttySSH]: https://www.ssh.com/academy/ssh/putty/windows/puttygen
[r2u]: https://github.com/eddelbuettel/r2u
[shiny]: https://shiny.posit.co/r/getstarted/
[sqlite]: https://sqlite.org/index.html
[systemd]: https://systemd.io/
[systemd_shiny]: https://hosting.analythium.io/using-systemd-to-run-shiny-apps/
[systemd_shiny_archive]: https://web.archive.org/web/20260819145512/https://hosting.analythium.io/using-systemd-to-run-shiny-apps/
