---
title: "intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lottie)
```

[`{lottie}`][lottie] is a R package with a [Shiny][shiny] app for capturing data from a long-running field study of Long
Tailed Tits in the Rivelin Valley on the edge of the city of Sheffield, UK.

## Installation

You can install the package directly from GitHub using [`{pak}`][pak], if you don't have `{pak}` installed install it first).

```shell
install.packages("pak")
pak::pkg_install("Morinay-lab/lotties")
```

## Launching

To launch the Shiny App locally you can run the following...

```R
shiny::runApp(system.file("shiny", package = "lotties"))
```

This should launch a browser pointing to a locally running instance of the application. The database will be in memory
and is _not_ permanent.

### Persistence

By default the App is set to run in memory and so no historical data nor data entered persists. When the application is
killed all entered data will be lost. Included in the package under `inst/xls` are various historical data files which
have been cleaned and added to a [SQLite][sqlite3] which resides under `inst/sqlite`. By modifying `testing <- FALSE` in
the `inst/shiny/server.R` file then the database with the historical data in the `inst/sqlite/` directory will be
loaded.

### Data Entry

There are a number of sections for data to be entered. It is expected that they will be completed in the order in which
they appear on the "accordian" on the left, i.e. _GPS > Observation Metadata > Flock Data > Individual Data >
Interactions Data_. Doing so means that the `Flock ID` entered under the _Flock Data_ will then be used to populate the
pull-down menus in subsequent sections (_Individual Data_ and _Interactions Data_).

## Deployment

To make the Shiny App available more widely the Morinay Lab have comissioned a Virtual Machine (VM) from University of
Sheffield IT Services.

This is a GNU/Linux machine running Ubuntu Linux and users can connect to the Shiny App that is running from any
computer on the University network or whilst connected to the Universities Virtual Private Network (VPN).

**TODO** Deploy the Shiny App on the server and work out the address to point to (https://lotties.shef.ac.uk/)


## Virtual Machine Administration

To connect to the virtual machine `lotties.shef.ac.uk` you will need to have an account created for you. For security
it has been configured to only allow connections using SSH authentication.

### Key Generation

#### GNU/Linux and OSX
If you are using GNU/Linux or OSX then you
can create SSH keys using the following commands (**NB** - It is recommend to use [Ed25519 Twisted Edward curve keys][ed25519])

```shell
ssh-keygen -o -a 100 -t ed25519
```

By default you should be prompted for a secure password.

#### Windows

Windows users should follow the instructions on [Using PuTTYgen on Windows to generate SSH key pairs][puttySSH] an

### Uploading Keys

Share your `id_25519.pub` (the public part of your key) with the administrator who will setup your account and place the
public key under your `/home/<username>/.ssh/id_ed25519.pub`

[ed25519]: https://blog.g3rt.nl/upgrade-your-ssh-keys.html
[pak]: https://pak.r-lib.org/
[lottie]: https://github.com/Morinay-lab/lotties
[puttySSH]: https://www.ssh.com/academy/ssh/putty/windows/puttygen
[shiny]: https://shiny.posit.co/r/getstarted/
[sqlite]: https://sqlite.org/index.html
