---
title: "SQLite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lottie)
```

This document provide a _very_ brief introduction to interacting with the SQLite database from R using
[RSQLite][rsqlite].

## Connection

You _always_ need to have a connection defined which points to the location of the database. This may be `:memory:` if
testing things locally in memory or it may be the location of a `.sql` file on disk.

```{r connection}
# In memory database
db_path <- ":memory:"
# On disk database
db_path <- "inst/sqlite/lottie.sql"
con <- DBI::dbConnect(RSQLite::SQLite(), db_path)
```

All subsequent commands for interacting with the database required the `con` (connection) so that they know what
database they are connecting to.

## Listing Tables

List tables with `RSQLite::dbListTables(con)`

```{r sql_list_tables}
> RSQLite::dbListTables(con)
[1] "Composition"  "Conditions"   "Description"  "GPS"          "Interactions"
[6] "OtherSpecies" "Person"       "Rings"        "Sites"
```

## Creating Tables

You create a dataframe in R and use this as the table to be written to the database. 

```{r sql_write_table}
df <- data.frame("user" = character(),
                 "date" = character(),
                 "start_time" = character(),
                 "end_time" = character(),
                 "weather" = character(),
                 "visibility" = character(),
                 "cloudy_grey" = logical(),
                 "foggy" = logical(),
                 "light_rain" = logical(),
                 "partly_cloudy" = logical(),
                 "really_rainy" = logical(),
                 "sunny" = logical(),
                 "windy" = logical(),
                 stringsAsFactors = FALSE,
                 check.names = FALSE)
RSQLite::dbWriteTable(conn = con,
                      name = "Conditions",
                      df,
                      overwrite = overwrite)

```

## Extract Data

You extract data by defining a query in SQL and then using the `RSQLite::dbGetQuery()` function. Typically the results
will then be stored in a dataframe.

### Complete table

You can extract a complete table very easily and convert the results to a [tibble][tibble] using the [`SELECT`][sqlite_select]. Here `*` is a wild card and
short hand for "all columns".

```{r sql_get_query_all}
table <- "Interactions"
query <- paste0("SELECT * FROM ", table)  ## NB - the space after FROM is essential
interactions_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

### Some columns

If you only want some columns from a table you have to be explicit as to which you want.

```{r sql_get_query_cols}
table <- "Interactions"
query <- paste0("SELECT date,time,flock_a,flock_b FROM ", table)  ## NB - the space after FROM is essential
interactions_subset_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

### Some rows

You may wish to filter the rows of a table which is done with the [`WHERE`][sqlite_where]. Here we select a specific date

```{r sql_get_query_rows}
table <- "Interactions"
query <- paste0("SELECT * FROM ", table, " WHERE date == '2026-08-20'")  ## NB - the space after FROM and before WHERE is essential
interactions_subset_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

## Modifying data

It is possible to modify data in a SQLite database using [`UPDATE`][sqlite_update]. You will need to know the
appropriate `WHERE` condition to select the row you wish to update and this will typically involve using other operators
such as
[`AND`][sqlite_and], [`OR`][sqlite_or], [`BETWEEN`][sqlite_between] or [`IN`][sqlite_in].

The example below flips round which flock was chasing which from `1` (`a`) chasing `4` (`b`) to `4` (`b`) chasing `1`
(`a`). The data

```{r sql_get_query_update}
table <- "Interactions"
query <- paste0("UPDATE ", ## NB - the space after FROM is essential
                table,
                " SET a_chasing_b = 0, b_chasing_a = 1",  ## NB - the spaces before keywords are also essential
                " WHERE date == '2026-08-20' ",
                " AND flock_a == '1' ",
                " AND flock_b == '4'",
                " AND a_chasing_b = 1",
                " AND b_chasing_a = 0")
RSQLite::dbSendQuery(con, query)
```


## Summary

This is just the tip of the iceberg there is a _lot_ to learn about writing database queries. A good resource for
learning more is the [SQLite Tutorial][sqlite_toturial].

[rsqlite]: https://rsqlite.r-dbi.org/
[sqlite_select]: https://www.sqlitetutorial.net/sqlite-select/
[sqlite_tutorial]: https://www.sqlitetutorial.net/
[sqlite_and]: https://www.sqlitetutorial.net/sqlite-and/
[sqlite_between]: https://www.sqlitetutorial.net/sqlite-between/
[sqlite_in]: https://www.sqlitetutorial.net/sqlite-in/
[sqlite_or]: https://www.sqlitetutorial.net/sqlite-or/
[sqlite_where]: https://www.sqlitetutorial.net/sqlite-where/
[sqlite_update]: https://www.sqlitetutorial.net/sqlite-update/
[tibble]: https://tibble.tidyverse.org/
