---
title: "SQLite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lottie)
```

This document provide a _very_ brief introduction to interacting with the SQLite database from R using
[RSQLite][rsqlite].

## Connection

You _always_ need to have a connection defined which points to the location of the database. This may be `:memory:` if
testing things locally in memory or it may be the location of a `.sql` file on disk.

```{r connection}
# In memory database
db_path <- ":memory:"
# On disk database
# The next two lines set the working directory, the firstr on the VM, the second on a users computer and should replace
# <username> with your username
setwd("/home/lotties/lotties")
# setwd("c:/Users/<username>/Documents")
db_path <- paste0(getwd(), "/inst/sqlite/lottie.sql")
con <- DBI::dbConnect(RSQLite::SQLite(), db_path)
```

All subsequent commands for interacting with the database required the `con` (connection) so that they know what
database they are connecting to.

## Listing Tables

List tables with `RSQLite::dbListTables(con)`

```{r sql_list_tables}
RSQLite::dbListTables(con)
#[1] "Composition"  "Conditions"   "Description"  "GPS"          "Interactions"
#[6] "OtherSpecies" "Person"       "Rings"        "Sites"
```

## Creating Tables

You create a dataframe in R and use this as the table to be written to the database.

```{r sql_write_table}
df <- data.frame("user" = character(),
                 "date" = character(),
                 "start_time" = character(),
                 "end_time" = character(),
                 "weather" = character(),
                 "visibility" = character(),
                 "cloudy_grey" = logical(),
                 "foggy" = logical(),
                 "light_rain" = logical(),
                 "partly_cloudy" = logical(),
                 "really_rainy" = logical(),
                 "sunny" = logical(),
                 "windy" = logical(),
                 stringsAsFactors = FALSE,
                 check.names = FALSE)
RSQLite::dbWriteTable(conn = con,
                      name = "Conditions",
                      df,
                      overwrite = overwrite)

```

## Extract Data

You extract data by defining a query in SQL and then using the `RSQLite::dbGetQuery()` function. Typically the results
will then be stored in a dataframe.

### Complete table

You can extract a complete table very easily and convert the results to a [tibble][tibble] using the [`SELECT`][sqlite_select]. Here `*` is a wild card and
short hand for "all columns".

```{r sql_get_query_all}
table <- "Interactions"
query <- paste0("SELECT * FROM ", table)  ## NB - the space after FROM is essential
interactions_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

### Some columns

If you only want some columns from a table you have to be explicit as to which you want.

```{r sql_get_query_cols}
table <- "Interactions"
query <- paste0("SELECT date,time,flock_a,flock_b FROM ", table)  ## NB - the space after FROM is essential
interactions_subset_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

### Some rows

You may wish to filter the rows of a table which is done with the [`WHERE`][sqlite_where]. Here we select a specific date

```{r sql_get_query_rows}
table <- "Interactions"
query <- paste0("SELECT * FROM ", table, " WHERE date == '2026-08-20'")  ## NB - the space after FROM and before WHERE is essential
interactions_subset_df <- RSQLite::dbGetQuery(con, query) |> dplyr::tibble()
```

## Modifying data

It is possible to modify data in a SQLite database using [`UPDATE`][sqlite_update]. You will need to know the
appropriate `WHERE` condition to select the row you wish to update and this will typically involve using other operators
such as
[`AND`][sqlite_and], [`OR`][sqlite_or], [`BETWEEN`][sqlite_between] or [`IN`][sqlite_in].

The example below flips round which flock was chasing which from `1` (`a`) chasing `4` (`b`) to `4` (`b`) chasing `1`
(`a`). The data

```{r sql_get_query_update}
table <- "Interactions"
query <- paste0("UPDATE ", ## NB - the space after UPDATE is essential
                table,
                " SET a_chasing_b = 0, b_chasing_a = 1",  ## NB - the spaces before keywords are also essential
                " WHERE date == '2026-08-20' ",
                " AND flock_a == '1' ",
                " AND flock_b == '4'",
                " AND a_chasing_b = 1",
                " AND b_chasing_a = 0")
RSQLite::dbSendQuery(con, query)
```

## ☣️Deleting Data ☣️

It is possible to also delete data from the SQLite database directory, but it is advisable to be wary when doing
so. Even though backups are made daily it is possible that some data may be lost.

It is recommended when attempting to delete data directly that you first construct a query that uses `SELECT` and
`WHERE` to identify the rows you think will be deleted and check that this is what you want to modify.

The [`DELETE`][sqlite_delete] command allows removal of one, multiple or all rows from a table. It takes an argument of
the `table` from which you wish to remove rows and a search condition using [`WHERE`][sqlite_where].

### Remove a one or more rows

First construct a [`SELECT`][sqlite_select] statement and see what is returned...

```{r sql_select_for_deletion}
table <- "Interactions"
select_query <- paste0("SELECT FROM ", ## NB - the space after FROM is essential
                table,
                " WHERE date == '2025-07-21' ",
                " AND user == 'LN' ")
RSQLite::dbSendQuery(con, select_query)
```

You may need to use multiple `AND` statements to narrow down the row(s) you want to remove. Once you are happy you have
identified just the rows you wish to remove you can then switch the query to use `DELETE` instead of `SELECT`

```{r sql_delete_single_row}
table <- "Interactions"
delete_query <- paste0("DELETE FROM ", ## NB - the space after FROM is essential
                table,
                " WHERE date == '2025-07-21' ",
                " AND user == 'LN' ")
RSQLite::dbSendQuery(con, delete_query)
```

You can re-run your `select_query` to check the row(s) have been deleted, it should not return any rows.

```{r sql_delete_single_row_check}
RSQLite::dbSendQuery(con, select_query)
<SQLiteResult>
  SQL  SELECT * FROM interactions WHERE date == '2025-07-21' AND user == 'LN'
  ROWS Fetched: 0 [complete]
       Changed: 0
Warning message:
Closing open result set, pending rows
```

### Remove all rows from a table

To remove all rows from a table you omit the `WHERE` clauses from the query.

```{r sql_delete_single_row}
table <- "Interactions"
delete_query <- paste0("DELETE FROM ", ## NB - the space after FROM is essential
                table)
RSQLite::dbSendQuery(con, delete_query)
```
## Restoring Data

Back-ups made daily to `.csv` files which are compressed to `.zip` files. It is possible to use these to restore the
database to the previous snapshot if required.

The location of backups is defined in the `.Renviron` and stored in the `LOTTIES_BACKUP_DIR`, at the time of writing
this was set to `/mnt/san/morinay_group/Shared/Database_Files/backup`.

To restore a backup you need to extract the files from the most recent `.zip`.

```shell
sudo su lotties
cd ~/lotties/tmp
mkdir $(date +%Y%m%d)     # Create a directory to extract files to of the form YYYYMMDD
unzip /mnt/san/morinay_group/Shared/Database_Files/backup/extract_2026-09-30_060003.847514.zip -d .
ls -lha 20260930
total 172K
drwxrwxr-x 2 lotties lotties 4.0K Sep 30 12:57 .
drwxrwxr-x 3 lotties lotties 4.0K Sep 30 12:56 ..
-rw-rw-r-- 1 lotties lotties 6.1K Sep 30 06:00 composition.csv
-rw-rw-r-- 1 lotties lotties  320 Sep 30 06:00 conditions.csv
-rw-rw-r-- 1 lotties lotties 1.2K Sep 30 06:00 description.csv
-rw-rw-r-- 1 lotties lotties 141K Sep 30 06:00 gps.csv
-rw-rw-r-- 1 lotties lotties  133 Sep 30 06:00 interactions.csv
```

Now change directory back to `~/lotties` so that all the relative paths will work and start R.

```shell
cd lotties
R
```

Setup a connection to the `inst/sqlite/lottie.sql` database and list the available tables.

```r
db_path <- paste0(getwd(), "/inst/sqlite/lottie-20260930.sql")
con <- DBI::dbConnect(RSQLite::SQLite(), db_path)
> RSQLite::dbListTables(con)
[1] "Composition"  "Conditions"   "Description"  "GPS"          "Interactions"
[6] "OtherSpecies" "Person"       "Rings"        "Sites"
```

We now load each of the `.csv` files and update the database. This **will** over-write anything that is in the database
as we use the option `overwite=TRUE` when we use the `RSQLite::dbWriteTable()`.

```r
## You may need to modify this line to reflect the current date
TMP_DIR <- "tmp/20260930"

## Composition
table <- "Composition"
composition_df <- read.csv(paste0(TMP_DIR, "composition.csv"))
RSQLite::dbWriteTable(conn = con, name=table, composition_df, overwrite=TRUE)

## Conditions
table <- "Conditions"
conditions_df <- read.csv(paste0(TMP_DIR, "conditions.csv"))
RSQLite::dbWriteTable(conn = con, name=table, conditions_df, overwrite=TRUE)

## Description
table <- "Description"
description_df <- read.csv(paste0(TMP_DIR, "description.csv"))
RSQLite::dbWriteTable(conn = con, name=table, description_df, overwrite=TRUE)

## GPS
table <- "GPS"
gps_df <- read.csv(paste0(TMP_DIR, "gps.csv"))
RSQLite::dbWriteTable(conn = con, name=table, gps_df, overwrite=TRUE)

## Interactions
table <- "Interactions"
interactions_df <- read.csv(paste0(TMP_DIR, "interactions.csv"))
RSQLite::dbWriteTable(conn = con, name=table, interactions_df, overwrite=TRUE)

```

## Summary

This is just the tip of the iceberg there is a _lot_ to learn about writing database queries. A good resource for
learning more is the [SQLite Tutorial][sqlite_toturial].

[rsqlite]: https://rsqlite.r-dbi.org/
[sqlite_select]: https://www.sqlitetutorial.net/sqlite-select/
[sqlite_tutorial]: https://www.sqlitetutorial.net/
[sqlite_and]: https://www.sqlitetutorial.net/sqlite-and/
[sqlite_between]: https://www.sqlitetutorial.net/sqlite-between/
[sqlite_delete]: https://www.sqlitetutorial.net/sqlite-delete
[sqlite_in]: https://www.sqlitetutorial.net/sqlite-in/
[sqlite_or]: https://www.sqlitetutorial.net/sqlite-or/
[sqlite_where]: https://www.sqlitetutorial.net/sqlite-where/
[sqlite_update]: https://www.sqlitetutorial.net/sqlite-update/
[tibble]: https://tibble.tidyverse.org/
